---
title: "Gov 50 Final Project"
author: "Jonah Simon"
description: "Gov 50 Final Project"
output:
  distill::distill_article:
    self_contained: false
---

## Project thoughts

I am interested in exploring data related to the filibuster and its impact on the amount of legislation passed by the Senate over time.

## Project proposal

Does the number of filibusters in the Senate impact Congressional approval ratings by the American public? In this study, I plan to analyze the relationship between cloture votes in the Senate and Congressional approval by year. My hypothesis is that an increase in the number of failed cloture votes will be accompanied by an decrease in public approval for Congress. This is because filibusters are often done to prevent legislation that a majority of Senators support but do not have the necessary supermajority of 60 votes. Thus, because Americans elected this Senate majority and thus a majority of Americans likely support the policies the majority party produces, I expect that they will disaprove of the use of the filibuster and that this will result in greater disaproval of Congress in general. For my data, I will use two data sets. The first data set comes from the Brookings Institute and tracks the number of cloture votes and the success of each vote from 1919-2018. The data is divided by Senate. The second comes from Gallup, and contains Congressional approval ratings from 1999-2023. Thus, the scope of my study will be the years 1999-2018, as this is where my two data sets overlap. Additionally, as the filibuster data set only tracks by Congress (a two year period) and not by year, my unit of analysis will be a Congress (e.g., the 118th Congress). I will use the average of the two years of approval data to compare to the filibuster data, as the approval data is separated by individual year and not by Congress. If I observe that there is a negative linear relationship between the number of filibusters and Congressional approval, this will provide support for my thesis. If I observe a positive linear relationship, or no significant relationship, this will provide evidence against my hypothesis.

## Data visualization

```{r LoadData}
library(tidyverse)
filibuster <- read.csv("Data/FilibusterData.csv")
congress <- read.csv("Data/CongressApproval.csv") # Loading data
```

```{r CleanData}
congress <- congress |>
  mutate(
    Session = if_else(Time%%2== 0, 2, 1)
  )

congress <- congress |>
  select(-X) # Cleaning Congress data 
```

```{r PrepData}
filibuster <- filibuster |>
  filter(Session == "1" | Session == "2")

filibuster <- filibuster |>
  mutate(
    Session = as.integer(Session)
  ) |>
  drop_na()

congress <- congress |>
  mutate(
    Session = as.integer(Session)
  ) |>
  drop_na() # Preparing data for merge
```


```{r MergeData}
data <- filibuster |>
  inner_join(congress)

data <- data |>
  select(-Years, -Session) |>
  rename(Year = Time) # Merging data
```

```{r FixPercentage}
data <- data |>
  mutate(`Congress Approval` = as.integer(sub("%", "", Approve))
  ) # Changing character to integer
```

```{r FilibusterVar}
data <- data |>
  mutate(Filibusters = Attempted - Successful) # Creating filibuster variable
```


```{r}
correlation_plot <- ggplot(data, aes(x = Filibusters, y = `Congress Approval`)) +
  geom_point(color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "indianred") +
  labs(
    title = "Impact of the Filibuster on Public Approval of Congress",
    subtitle = "Using data from Gallup and The Brookings Institute",
    x = "Number of Filibusters In Senate",
    y = "% Public Approval of Congress"
  )

correlation_plot # Plotting data
```

## Analysis

```{r}
regression <- lm(`Congress Approval` ~ Filibusters, data = data)
library(broom)

regression |>
  broom::tidy() |>
  select(term, estimate) |>  
  knitr::kable(digits = 2)
```

In this analysis, I ran a regression of the relationship between the number of filibusters in the Senate (measured as the difference between cloture votes taken and successful cloture votes), and public approval of Congress. My findings are that the intercept is 34.17. This means that if there were zero filibusters in the Senate, the estimated public approval of the Senate would be 34.17%. The correlation between filibusters and public approval is -0.33. This means that for each filibuster that occurs in the Senate, public approval of Congress decreases by .33 percentage points. 

```{r}
ggplot(data, aes(x = Year, y = Filibusters)) +
  geom_col(fill = "dodgerblue4") +
  labs(
    title = "Senate Filibusters per Year (1999-2018)"
  )
```

I also developed a bar plot to demonstrate the number of filibusters occurring each year in the Senate during the time of my analysis from 1999-2018. There does not appear to be any significant trend over time.

## Introduction

In this project, I examine the relationship between filibusters in the U.S. Senate and public approval of the U.S. Congress from the years 1999-2018. The Senate rules require a supermajority of 60 votes to pass a cloture motion, which is used to terminate debate on a bill and allow for the bill to be brought to a floor vote. A filibuster is a term that refers to a failed cloture vote. Filibusters are a tactic commonly used by the minority party to block legislation that would otherwise pass a floor vote by simple majority. Thus, the number of filibusters that occur in a year are a strong proxy for the degree of Congressional gridlock, which is a more abstract concept. 

I chose to examine the relationship between filibusters and public approval of Congress in order to better understand the degree to which the actual performance of Congress relates to the public's perception of the institution. My hypothesis is that the number of filibusters occurring in a year will be negatively correlated with public approval of Congress. This is because the filibuster is often used to prevent the passage of legislation that the majority of the Senate, and thus presumably a majority of the public, supports. However, I also hypothesize that the relationship will be weak, as I expect that a large portion of the public will be uninterested and uninformed on the events of Congress and will have fairly stable opinions on its performance across time.

## Data

In this study, I conduct a cross-sectional analysis drawing from two different data sets. For my data on the number of filibusters each year, I used a data set from the Brookings Institute which tracks cloture votes in the Senate from 1919-2018 using data from the Congressional record. To find the number of filibusters that occurred each year, I took the difference of the total number of cloture votes and the number of successful cloture votes. For my data on Congressional approval, I utilized a Gallup survey taken each year from 1999-2023 tracking Congressional approval among Americans using random sampling. Combining the two data sources, I was limited in the scope of my analysis to the years 1999-2018, as these were the years at which the data from the two sources overlapped.

Figure 1 shows the trend of the number of filibusters occurring over the period of anaylsis. Figure 2 shows the trend of Congressional approval over the period of anaysis.

```{r}
ggplot(data, aes(x = Year, y = Filibusters)) +
  geom_col(fill = "dodgerblue4") +
  labs(
    title = "Senate Filibusters per Year (1999-2018)",
    subtitle = "Using Brookings Institute Data"
  )
```
Fig. 1

```{r}
ggplot(data, aes(x = Year, y = `Congress Approval`)) +
  geom_line(color = "dodgerblue4", size = 1) +
  labs(
    title = "% Public Approval of Congress (1999-2018)",
    subtitle = "Using Gallup Survey Data",
    y = "% Public Approval of Congress"
  )
```

Fig. 2

## Results

```{r}
correlation_plot
```

Fig. 3

```{r}
regression |>
  broom::tidy() |>
  select(term, estimate) |>  
  knitr::kable(digits = 2)
```

